Human capital mediates natural selection in contemporary humans

### 2022-05-04 why-natural-selection.Rmd
Many tweaks to text. Weight mediation bootstraps.

---
title: "Human capital mediates natural selection in contemporary humans"
author: "David Hugh-Jones \\thanks{Corresponding author. School of Economics, University of East Anglia, Norwich, UK. Email: D.Hugh-Jones@uea.ac.uk}, Abdel Abdellaoui \\thanks{Department of Psychiatry, Amsterdam UMC, University of Amsterdam, Amsterdam, The Netherlands. Email: a.abdellaoui@amsterdamumc.nl}"
date: "`r format(Sys.Date(), '%d %B %Y')`"
abstract: "\\noindent Natural selection has been documented in contemporary humans, but  little is known about the mechanisms behind it. We test for natural selection through the association between 33 polygenic scores and fertility, across two generations, using data from UK Biobank (N = 409,629 British subjects with European ancestry). Consistently over time, polygenic scores that predict lower earnings, education and health also predict lower fertility. Selection effects are concentrated among lower SES groups, younger parents, people with more lifetime sexual partners, and people not living with a partner. The direction of natural selection is reversed among older parents, or after controlling for age at first live birth. These patterns are in line with economic theories of fertility, in which earnings-increasing human capital may either increase or decrease fertility via income and substitution effects in the labour market. Studying natural selection can help us understand the genetic architecture of health outcomes: we find evidence in modern day Great Britain for multiple natural selection pressures that vary between subgroups in the direction and strength of their effects, that are strongly related to the socio-economic system, and that may contribute to health inequalities across income groups."
bibliography: "why-natural-selection.bib"
output: 
  bookdown::pdf_document2:
    toc: false
    latex_engine: xelatex
    number_sections: true
    keep_tex: true
header-includes:
  - \usepackage{subfig}
  - \usepackage{setspace}\doublespacing
  - \usepackage{placeins}
  - \usepackage[format=plain, labelfont={bf,it}, textfont=it]{caption}
  - \usepackage{titlesec}
  - \titleformat*{\section}{\sffamily\LARGE}
  - \titleformat*{\subsection}{\sffamily\itshape\Large}
  - \hypersetup{colorlinks = true, linkcolor = {blue}, linkbordercolor = {white}}
  - \usepackage{amsthm}
  - \theoremstyle{plain}
  - \newtheorem{lem}{\protect\lemmaname}
  - \providecommand{\lemmaname}{Lemma}
  - \usepackage{etoc}
  - \usepackage{lineno}
  - \linenumbers
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
mainfont: Baskerville
mathfont: Baskerville
sansfont: "Gill Sans"
---

@@

  dfr %>% mutate(
    !! score_col := pretty_names(!! score_col),
    !! score_col := fct_reorder(!! score_col, estimate, order_abs(order_idx))
  ) %>% 
    rename(p = p.value) %>% 
    ggplot(aes(estimate, {{score_col}}, yend = {{score_col}}, 
               shape = {{fill_col}}, ...)) +
    my_vline + 
    confint_geom_segment +
    geom_point(size = 1.8, aes(color = {{fill_col}})) +
    geom_point(size = 0.7, color = "white", aes(alpha = p < 0.05/33)) +
    my_fill_scale(aesthetics = "color", n = n_cats, direction = fill_direction) +
    scale_shape_manual(values = 1:n_cats + 14) +
    my_labs() +
    my_theme() +
    scale_linetype_manual(name = "", values = c("95% c.i." = 1)) +
    scale_alpha_discrete(range = c(1, 0),
                         guide = guide_legend(
                           override.aes = list(
                             shape = c(1, 16),
                             color = "black",
                             alpha = 1,
                             size = 1.8
                           )
                         ))
}

@@

pval_geom_point <- function (size = 1.6, stroke = 0.8, fill = NULL, ...) {
  if (is.null(fill)) {
    geom_point(shape = 21, size = size, stroke = stroke, ...)
  } else {
    geom_point(shape = 21, size = size, stroke = stroke, fill = fill, ...)
  }
}

@@

```{r plot-means-over-time, fig.height = 6, fig.cap = "Mean polygenic scores (PGS) by birth year in UK Biobank. Circles are means for 5-year intervals. Lines are 95% confidence intervals. Solid circles show a significant linear increase (green) or decrease (red) over time (p < 0.05/33)."}

@@

pgs_over_time %>% 
      mutate(score_name = pretty_names(score_name)) %>% 
      ggplot(aes(YOB, group = score_name, colour = Change, shape = Change)) + 
      geom_linerange(aes(ymin = score_lo, ymax = score_hi), color = "grey60", 
                       size = 3) +
      geom_point(aes(y = score), size = 2) +
      facet_wrap(~score_name, ncol = 6) +
      my_hline +
      scale_color_manual(values = c("-" = "darkred", "+" = "darkgreen", "o" = "black")) +
      scale_shape_manual(values = c("-" = "circle", "+" = "circle", "o" = "circle open")) +
      theme(
        line        = element_line(size = 0.5, lineend = "round"),
        strip.text  = element_text(size = 6), 
        axis.text.x = element_text(size = 7),
        legend.position = "none"
      ) +
      labs(y = "PGS", x = "Birth year (5 year intervals)")

@@

To test more directly for natural selection, we regress respondents' relative 
lifetime reproductive success (RLRS) on their polygenic scores (PGS):

@@

Figure \@ref(fig:plot-res-wt) plots selection effects in the entire
sample.[^balance-diverse-def] To correct for ascertainment bias, we weight
participants using the model of @van2021effects. Weighting makes a
large difference: effect sizes go up by a mean of 
`r percent(mean_rel$Main - 1)`.[^weighting] `n_sig` out of 33 selection effects
are significant at $p$ < 0.05/33.

@@

res_wt_combined %>% 
      filter(Weights %in% c("Main", "None")) %>% 
      mutate(
        Weights  = recode(Weights, "Main" = "Weighted", "None" = "Unweighted"),
        term     = pretty_names(term),
        term     = fct_reorder(term, estimate, order_abs(1))
      ) %>% 
      rename(p = p.value) %>% 
      # can't use standard_ggplot because we have non-standard geom_point()
      ggplot(aes(
        x      = estimate, 
        y      = term, 
        yend   = term,
        shape  = Weights, 
        fill   = Weights, 
        colour = p < 0.05/{{n_regs}}
      )) +
      my_vline +
      geom_point(size = 1.6, stroke = 0.8, alpha = 0.9) +
      confint_geom_segment() +
      pval_scale() +
      scale_fill_manual(values = c(
        "Unweighted"        = "grey15", 
        "Weighted"          = "darkseagreen4"
      )) +
      scale_shape_manual(values = c(
        "Unweighted" = "square filled", 
        "Weighted"   = "circle filled"
      )) + 
      labs(x = "Effect size", y = "") +
      scale_linetype_manual(name = "", values = c("95% c.i. uncorrected" = 1)) +
      my_theme()

@@

We now show the empirical puzzles which motivate our economic model. Each
concerns differences in the strength of natural selection across different
subgroups in the sample. Splitting the sample by demographic and social
variables, we re-estimate \@ref(eq:regression) within each subgroup. We use
socio-economic variables including income and education, and family structure
variables including age at first live birth, presence of a partner, and lifetime
number of sexual partners.

Figure \@ref(fig:plot-income-educ-level) plots selection effects for each
polygenic score, grouping respondents by age of completing full-time education,
and by household income. Effects are larger and more significant for the lowest
income category, and for the lowest education category. These results are robust
to controlling for respondents' age (Appendix section \@ref(sec:age-control)).
Turning to family structure, we split respondents by lifetime number of sexual
partners, at the median value of 3 (Figure \@ref(fig:plot-n-partners)a).
Now, selection effects are larger and more significant among
those with more than 3 lifetime partners. Next we split respondents by whether
they were living with a spouse or partner at the time of interview.
Effects are larger among those not living with a spouse or partner
(Figure \@ref(fig:plot-n-partners)b).[^sex]

@@

We emphasize that our categories are not exogenous to polygenic scores. For
example -- both in the data (Appendix Figure \@ref(fig:plot-age-flb-dv)) and in
our theoretical model -- age at first live birth is a choice variable, which is
endogenous to a person's human capital and their related polygenic scores.
Nevertheless, differences in selection effects across subgroups constrain the
set of possible theories; a good theory of contemporary natural selection needs
to explain how these results come about. As we show below, a model based on the
economic theory of fertility can do just that.

We also examine selection effects among respondents' parents, using information
on respondents' number of siblings to calculate parents' RLRS. Effect sizes of
polygenic scores are highly correlated across the two generations (Appendix
Figure \@ref(fig:plot-res-sibs)). Median-splitting respondents by year of birth,
we find little evidence of change in effect sizes among the parents' generation.
There is some evidence that selection effect sizes are increasing in the 
respondents' generation, with 9 polygenic scores showing a significant increase. 
We also check whether selection effects vary by
socio-economic status and AFLB in the parents' generation. We have no
information about parents' income, so we use the 1971 Townsend deprivation score
of respondents' birthplace as a proxy [@townsend1987deprivation]. Results show
the same pattern as for respondents. Effect sizes are larger and more often
significant in the most deprived areas (Appendix Figure
\@ref(fig:plot-siblings-townsend)). Effects are larger among younger parents,
and change sign when controlling for AFLB, among both mothers and fathers
(Appendix Figures \@ref(fig:plot-age-birth-parents-cross), 
\@ref(fig:plot-age-birth-parents)). Lastly, we check for a
"quantity-quality tradeoff" between parents' number of children and number of
grandchildren. We don't find any: in fact, the correlation between respondents' 
and parents' RLRS is positive ($\rho$ = `r sib_chn_cor$estimate`, 
$p < 2 \times 10^{-16}$).

@@

To show this, we write down a simple model of fertility choices. $h$ is
an individual's level of human capital. For now, we simply identify this with
his or her wage $W$. Raising a child takes time $b$. People maximize utility $U$
from the number of children $N$ and from income $Y\equiv(1-bN)W$:

\[
U = u(Y)+aN.
\]

Here $a$ captures the strength of preference for children. $u(\cdot)$ captures
the taste for income, and is increasing and concave. We treat $N$ as continuous,
in line with the literature: this can be thought of as the expected number of
children among people with a given $a$, $b$ and $W$. The marginal
benefit of an extra child is $\frac{dU}{dN} = -bWu'(Y)+a$. The effect of an
increase in human capital on this marginal benefit is

\[
\frac{d^{2}U}{dNdh}=\underbrace{-bu'(Y)}_{\textrm{Substitution effect}}\underbrace{-bYu''(Y)}_{\textrm{Income effect}}.
\]

The *substitution effect* is negative and reflects that when wages increase,
time devoted to childcare costs more in foregone income. The positive *income
effect* depends on the curvature of the utility function, and reflects that when
income is higher, the marginal loss of income from children is less painful.

To examine education and fertility timing, we extend the model to
two periods. For convenience we ignore time discounting, and assume
that credit markets are imperfect so that agents cannot borrow. Write

\begin{equation}
U(N_{1},N_{2}) = u(Y_{1}) + u(Y_{2}) + aN_{1} + aN_{2}\label{eq:U}
\end{equation}

Instead of identifying human capital with wages, we now allow individuals to
choose a level of education $s \in [0,1]$, which has a time cost in period 1.
Education is complementary to human capital $h > 0$, and increases period 2
wages, which take the  simple functional form $w(s,h) = sh$. We normalize period
1 wages to 1, and let $u(\cdot)$ take the constant relative risk aversion form
$u(y)=\frac{y^{1 - \sigma} - 1}{1 - \sigma}$. $\sigma > 0$ measures the
curvature of the utility function, i.e. the decline in marginal utility of
income as income increases. We examine total fertility 
$N^{*} = N_{1}^{*} + N_{2}^{*}$ and the *fertility-human capital relationship*,
$\frac{dN^{*}}{dh}$. For $\sigma < 1$ in a neighbourhood of $\sigma = 1$, Table
\@ref(tab:theory) shows five theoretical predictions, along with our
corresponding empirical results for the correlation between polygenic scores and
RLRS.[^one-period-results] The key insight of the model is that for middling
levels of $\sigma$, the substitution effect dominates at low income levels, but
is balanced by the income effect as income increases.

@@

Thus, a simple economic model can explain many of our results.
Other empirical work in economics also supports our mechanism.
@caucutt2002women and @monstad2008education show that education and skills
affect age at first birth and fertility. Income decreases fertility at low
income levels, but increases it at higher income levels [@cohen2013financial].
US fertility decreases faster with education among single mothers than married
mothers [@baudin2015fertility], in line with our prediction 3 and as predicted by
@becker1981treatise. A related literature shows negative correlations
between IQ and fertility [e.g. @lynn2004new; @reeve2018systematic].

@@

We test this explanation in two ways. First, the economic theory predicts that
genetic variants will be selected for (or against) in proportion to their
correlation with human capital. Figure \@ref(fig:plot-cors-earnings-educ) plots
selection effects on each polygenic score against that score's correlation with
two measures of human capital: earnings in a respondent's first job, and
educational attainment. The relationships are strongly negative. Thus, human
capital appears to be relevant to natural selection. The negative relationship
suggests that substitution effects dominate income effects, which fits the known
negative association between income and fertility
[@becker1960economic;@jones2006economic]. The correlations reverse when we
control for age at first live birth, suggesting that within AFLB categories, the
income effect dominates.

@@

Second, for the `r nrow(res_mediation)` scores where the bivariate correlation in
\eqref{eq:regression} is significant at $p < 0.05/33$, we run a mediation
analysis to directly test whether the correlation between each polygenic score
and fertility is mediated by educational attainment (Appendix Table
\@ref(tab:tbl-res-mediation)). Figure \@ref(fig:plot-mediation) shows estimated
proportions explained along with bootstrap 95% confidence intervals
(uncorrected; 100 bootstraps). For `r n_pos_sig` scores, the indirect effect of
the score on fertility *via* educational attainment takes the same sign as the
overall effect, and is significantly different from zero ($p < 0.05/33$). Among
these scores, the median proportion of the total effect explained by the
indirect effect is `r percent(median_prop)`. The educational attainment variable
is a relatively crude measure of human capital: more accurate measures would
likely explain more of the total effect.

```{r plot-mediation, fig.align = "center", fig.cap = "Proportion of selection effect mediated by educational attainment, polygenic scores with significant selection effects"}

res_mediation %>% 
      mutate(
        term = pretty_names(term),
        term = fct_reorder(term, prop_ind, order_abs(1)),
        `Indirect effect p < 0.05/33` = sig 
      ) %>% 
      ggplot(aes(y = term)) + 
        geom_col(aes(x = prop_ind, fill = `Indirect effect p < 0.05/33`), 
                   width = .5) + 
        geom_errorbar(aes(
                          xmin = prop_ind_conf_low, 
                          xmax = prop_ind_conf_high,
                          linetype = "Bootstrap 95% c.i.",
                        ),
                        width = 0.3,
                        color = "grey50"
                      ) +
        scale_fill_manual(values = pval_palette) +
        my_vline +
        scale_x_continuous(labels = scales::percent, n.breaks = 7) +
        coord_cartesian(xlim = c(-0.5, 1)) +
        labs(x = "Proportion mediated", y = "") +
        scale_linetype_manual(name = "", values = c("Bootstrap 95% c.i." = 1))

@@

We consider three alternative theories that might explain our results. First,
welfare benefits which incentivize child-bearing might be taken up more among
low-income individuals. However, the majority of effect sizes appear unchanged
over a large span of twentieth-century history (Appendix Table
\@ref(tab:tbl-change-by-period)), during
which government spending on child-related benefits varied considerably
[@hoc1999socsec]. In general, there is only weak evidence that welfare benefits
affect fertility [@gauthier2007impact]. Future
work could test this theory more explicitly. A second alternative theory is that
polygenic scores correlate with the motivation to have children, i.e. parameter
$a$ in the model [cf. @jones2008fertility]. This theory would not explain why
selection effects are smaller at higher incomes and education levels. Indeed,
in the model, the effect of $a$ on fertility becomes stronger at higher levels
of human capital. A third alternative is that traits under selection are linked
to externalizing behaviour and risk-seeking. This might be partially captured by
our parameter $\sigma$, which can be interpreted as a measure of risk aversion
over income. However, a more direct channel is risky sexual behaviour
[@mills2021identification]. The data here provide some support for this: scores
which might plausibly be linked to externalizing behaviour, like ADHD and
younger age at smoking initiation, are selected for. This theory is less good,
though, at explaining variation in selection across the full range of scores,
including physical measures such as waist-hip ratio and BMI. We test the theory
directly by re-estimating equation \@ref(eq:regression) controlling for a
measure of risk attitude (UK Biobank field 2040).  The median ratio of effect
sizes between regressions with and without controls is `r median(ratio)`; all
scores which are significant at $p < 0.05/33$ in uncontrolled regressions remain
so when controlling for risk attitude. This non-result could simply reflect the
imprecision of the risk attitude measure, which is a single yes/no question.
But this measure *does* predict the overall number of children, highly
significantly ($p < 2 \times 10^{-16}$ in `r n_risk_sig` out of 33 regressions).
Given that, and the statistical power we get from our sample size, we believe
that the non-result is real: while risk attitude does predict fertility in the
sample, it is not an important channel for natural selection.

@@

To learn more about this, we compare effect sizes excluding and including
childless people in the *current* generation. The correlation between the two
sets of effect sizes is `r cor_inc_vs_excl_childless`. So, patterns across
different scores are broadly similar whether the childless are counted or not.
However, absolute effect sizes are smaller when the childless are excluded, for
`r n_smaller` out of 33 scores; the median percentage change is 
`r percent(median_pct_diff)`.

>>>>

---
title: "Human capital mediates natural selection in contemporary humans"
author: "David Hugh-Jones \\thanks{Corresponding author. School of Economics, University of East Anglia, Norwich, UK. Email: D.Hugh-Jones@uea.ac.uk}, Abdel Abdellaoui \\thanks{Department of Psychiatry, Amsterdam UMC, University of Amsterdam, Amsterdam, The Netherlands. Email: a.abdellaoui@amsterdamumc.nl}"
date: "`r format(Sys.Date(), '%d %B %Y')`"
abstract: "\\noindent Natural selection has been documented in contemporary humans, but  little is known about the mechanisms behind it. We test for natural selection through the association between 33 polygenic scores and fertility, across two generations, using data from UK Biobank (N = 409,629 British subjects with European ancestry). Consistently over time, polygenic scores that predict lower earnings, education and health also predict lower fertility. Selection effects are concentrated among lower SES groups, younger parents, people with more lifetime sexual partners, and people not living with a partner. The direction of natural selection is reversed among older parents, or after controlling for age at first live birth. These patterns are in line with the economic theory of fertility, in which earnings-increasing human capital may either increase or decrease fertility via income and substitution effects in the labour market. Studying natural selection can help us understand the genetic architecture of health outcomes: we find evidence in modern day Great Britain for multiple natural selection pressures that vary between subgroups in the direction and strength of their effects, that are strongly related to the socio-economic system, and that may contribute to health inequalities across income groups."
bibliography: "why-natural-selection.bib"
output: 
  bookdown::pdf_document2:
    toc: false
    latex_engine: xelatex
    number_sections: true
    keep_tex: true
header-includes:
  - \usepackage{subfig}
  - \usepackage{setspace}\doublespacing
  - \usepackage{placeins}
  - \usepackage[format=plain, labelfont={bf,it}, textfont=it]{caption}
  - \usepackage{titlesec}
  - \titleformat*{\section}{\sffamily\LARGE}
  - \titleformat*{\subsection}{\sffamily\itshape\Large}
  - \hypersetup{colorlinks = true, linkcolor = {blue}, linkbordercolor = {white}}
  - \usepackage{amsthm}
  - \theoremstyle{plain}
  - \newtheorem{lem}{\protect\lemmaname}
  - \providecommand{\lemmaname}{Lemma}
  - \usepackage{etoc}
  - \usepackage{lineno}
  - \linenumbers
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
mainfont: Baskerville
mathfont: Baskerville
sansfont: "Gill Sans"
---

@@

  confint_geom_segment <- if (conf_int) confint_geom_segment() else NULL

  dfr %>% mutate(
          !! score_col := pretty_names(!! score_col),
          !! score_col := fct_reorder(!! score_col, estimate, order_abs(order_idx))
        ) %>% 
        rename(p = p.value) %>% 
        ggplot(aes(estimate, {{score_col}}, yend = {{score_col}}, 
                     fill = {{fill_col}}, ..., colour = p < 0.05/{{n_regs}})) +
          my_vline + 
          confint_geom_segment +
          pval_geom_point(fill = if(rlang::quo_is_null(fill_col)) "steelblue") +
          pval_scale() +
          my_fill_scale(n = n_cats, direction = fill_direction) +
          my_labs() +
          my_theme() +
          scale_linetype_manual(name = "", 
                                  values = c("95% c.i. uncorrected" = 1)) +
          guides(
            fill     = guide_legend(order = 1),
            stroke   = guide_legend(order = 2),
            linetype = guide_legend(order = 3)
          )
}

@@

pval_geom_point <- function (size = 1.6, stroke = 0.9, fill = NULL, ...) {
  if (is.null(fill)) {
    geom_point(shape = 21, size = size, stroke = stroke, ...)
  } else {
    geom_point(shape = 21, size = size, stroke = stroke, fill = fill, ...)
  }
}

@@

```{r plot-means-over-time, fig.height = 7, fig.cap = "Mean polygenic scores (PGS) by birth year in UK Biobank. Circles are means for 5-year intervals. Lines are 95% confidence intervals. Triangles show a significant linear increase or decrease over time (p < 0.05/33)."}

@@

pgs_over_time %>% 
      mutate(score_name = pretty_names(score_name)) %>% 
      ggplot(aes(YOB, group = score_name, colour = Change, shape = Change)) + 
      geom_linerange(aes(ymin = score_lo, ymax = score_hi), color = "grey60", 
                       size = 3) +
      geom_point(aes(y = score), size = 1.5) +
      facet_wrap(~score_name, ncol = 5) +
      my_hline +
      scale_color_manual(values = c("-" = "darkred", "+" = "darkgreen", "o" = "black")) +
      scale_shape_manual(values = c(
                                      "-" = "triangle down open", 
                                      "+" = "triangle open", 
                                      "o" = "circle open"
                                    )) +
      theme(
        line        = element_line(size = 0.5, lineend = "round"),
        strip.text  = element_text(size = 6), 
        axis.text.x = element_text(size = 7),
        legend.position = "none"
      ) +
      labs(y = "PGS", x = "Birth year (5 year intervals)")

@@

To test for natural selection more directly, we regress respondents' relative 
lifetime reproductive success (RLRS) on each polygenic score (PGS):

@@

Figure \@ref(fig:plot-res-wt) plots selection effects in the whole
sample.[^balance-diverse-def] To correct for ascertainment bias, we use
participant weights from @van2021effects, which match the UK Biobank eligible
population on sex, birth year, location, education, employment, health,
household size and tenure, number of cars and age at death. Weighting makes a
large difference: effect sizes go up by a mean of 
`r scales::percent(mean_rel$Main - 1)`.[^weighting] `r n_sig` out of 33 weighted
selection effects are significant at $p$ < 0.05/33.

@@

res_wt_combined %>% 
      filter(Weights %in% c("Main", "None")) %>% 
      mutate(
        Weights  = recode(Weights, "Main" = "Weighted", "None" = "Unweighted"),
        term     = pretty_names(term),
        term     = fct_reorder(term, estimate, order_abs(1))
      ) %>% 
      standard_ggplot(fill_col = Weights, n_regs = n_regs)

@@

We now show the empirical puzzles which motivate our economic model. Each
concerns differences in the strength of natural selection across different
subgroups in the sample. We re-estimate \@ref(eq:regression) splitting the
sample by demographic and social variables, including income and education, and
family structure variables including age at first live birth, presence of a
partner, and lifetime number of sexual partners.

```{r calc-pct-diffs}
drake::loadd(res_edu)
drake::loadd(res_income)
drake::loadd(res_partners_joint)
drake::loadd(res_with_partner)
res_partners_joint %<>% filter(grepl(":", term))
res_with_partner %<>% filter(grepl(":", term))

res_edu_wide <- full_join(
                  res_edu %>% filter(age_fte_cat == "< 16"),
                  res_edu %>% filter(age_fte_cat == "> 18"),
                  by = "score_name"
                )

res_income_wide <- full_join(
                     res_income %>% filter(income_cat == 1),
                     res_income %>% filter(income_cat == 5),
                     by = "score_name"
                   )

res_partners_wide <- full_join(
                       res_partners_joint %>% filter(grepl("lo_partnersFALSE", term)),
                       res_partners_joint %>% filter(grepl("lo_partnersTRUE", term)),
                       by = "score_name"
                     )

res_with_partner_wide <- full_join(
                           res_with_partner %>% 
                             filter(grepl("with_partnerFALSE", term)),
                           res_with_partner %>% 
                             filter(grepl("with_partnerTRUE", term)),
                           by = "score_name"
                         )

mean_pct_diff <- function (dfr) {
   prop <- dfr %>% filter(
                    p.value.x < 0.05/33,
                    sign(estimate.x) == sign(estimate.y)
                  ) %>% 
                  mutate(
                    pct_diff = estimate.x/estimate.y
                  ) %>% 
                  pull(pct_diff) %>% 
                  mean()
   # percentage difference:
   prop - 1
}

pct_diff_edu <- mean_pct_diff(res_edu_wide)
pct_diff_income <- mean_pct_diff(res_income_wide)
pct_diff_n_partners <- mean_pct_diff(res_partners_wide)
pct_diff_with_partner <- mean_pct_diff(res_with_partner_wide)
```

Figure \@ref(fig:plot-income-educ-level) plots selection effects for each
polygenic score, grouping respondents by age of completing full-time education,
and by household income. Effects are larger and more significant for the lowest
education category, and for the lowest income category. The mean percentage
difference between the lowest and highest education categories, among scores
which are significant for the lowest category and have the same sign across
categories, is `r scales::percent(pct_diff_edu)`. Between the lowest and
highest income categories, it is `r scales::percent(pct_diff_income)`. These
results are robust to controlling for respondents' age (Appendix section
\@ref(sec:age-control)). Turning to family structure, we split respondents by
lifetime number of sexual partners, at the median value of 3 (Figure
\@ref(fig:plot-n-partners)a). Now, selection effects are larger and more
significant among those with more than 3 lifetime partners, with a mean percentage
difference of `r scales::percent(pct_diff_n_partners)`. Next we split
respondents by whether they were living with a spouse or partner at the time of
interview (Figure \@ref(fig:plot-n-partners)b).  Effects are larger among those not living with a spouse or partner. The mean percentage difference is 
`r scales::percent(pct_diff_with_partner)`.[^sex]

@@

We emphasize that these categories are not exogenous to polygenic scores. For
example -- both in the data (Appendix Figure \@ref(fig:plot-age-flb-dv)) and in
our theoretical model -- education and age at first live birth are choice
variables, which are endogenous to a person's human capital and to relevant
polygenic scores. Nevertheless, differences in selection effects across
subgroups constrain the set of possible explanations. A good theory of contemporary
natural selection needs to show how these differences come about. As we describe
below, a model based on the economic theory of fertility can do just that.

We also examine selection effects among respondents' parents, using information
on respondents' number of siblings to calculate parents' RLRS. Effect sizes of
polygenic scores are highly correlated across the two generations (Appendix
Figure \@ref(fig:plot-res-sibs)). Median-splitting respondents by year of birth,
we find little evidence of change in effect sizes among the parents' generation.
There is some evidence that selection effect sizes are increasing in the 
respondents' generation, with 9 polygenic scores showing a significant increase. 
We also check whether selection effects vary by
AFLB and socio-economic status in the parents' generation, using the 1971 
Townsend deprivation score of respondents' birthplace as a proxy for income
[@townsend1987deprivation]. Results show the same pattern as for the
respondents' generation. Effect sizes are larger and more often significant in
the most deprived areas (Appendix Figure \@ref(fig:plot-siblings-townsend)).
Effects are larger among younger fathers and mothers, and change sign when
controlling for AFLB (Appendix Figures \@ref(fig:plot-age-birth-parents-cross),
\@ref(fig:plot-age-birth-parents)). Lastly, we check for a "quantity-quality
tradeoff" between parents' number of children and number of grandchildren. We
don't find any: in fact, the correlation between respondents' and parents' RLRS
is positive ($\rho$ = `r sib_chn_cor$estimate`, $p < 2 \times 10^{-16}$).

@@

To show this, consider a simple model of fertility choices. $h$ is
an individual's level of human capital. For now, we simply identify this with
his or her wage $W$. Raising a child takes time $b$. People maximize utility $U$
from the number of children $N$ and from income $Y\equiv(1-bN)W$:
\[
U = u(Y)+aN.
\]
Here $a$ captures the strength of preference for children. $u(\cdot)$ captures
the taste for income, and is increasing and concave. We treat $N$ as continuous,
in line with the literature: this can be thought of as the expected number of
children among people with a given $a$, $b$ and $W$. The marginal
benefit of an extra child is $\frac{dU}{dN} = -bWu'(Y)+a$. The effect of an
increase in human capital on this marginal benefit is
\[
\frac{d^{2}U}{dNdW}=\underbrace{-bu'(Y)}_{\textrm{Substitution effect}}\underbrace{-bYu''(Y)}_{\textrm{Income effect}}.
\]
The *substitution effect* is negative and reflects that when wages increase,
time devoted to childcare costs more in foregone income. The positive *income
effect* depends on the curvature of the utility function, and reflects that when
income is higher, the marginal loss of income from children is less painful.

To examine education and fertility timing, we extend the model to
two periods. For convenience we ignore time discounting, and assume
that credit markets are imperfect so that agents cannot borrow. Write
\begin{equation}
U(N_{1},N_{2}) = u(Y_{1}) + u(Y_{2}) + aN_{1} + aN_{2}\label{eq:U}
\end{equation}
Instead of identifying human capital with wages, we now allow individuals to
spend time $s \in [0,1]$ on education in period 1.
Education is complementary to human capital $h > 0$, and increases period 2
wages, which take the simple functional form $w(s,h) = sh$. We normalize period
1 wages to 1, and let $u(\cdot)$ take the constant relative risk aversion form
$u(y)=\frac{y^{1 - \sigma} - 1}{1 - \sigma}$. $\sigma > 0$ measures the
curvature of the utility function, i.e. the decline in marginal utility of
income as income increases. We examine total fertility 
$N^{*} = N_{1}^{*} + N_{2}^{*}$ and the *fertility-human capital relationship*,
$\frac{dN^{*}}{dh}$. For $\sigma < 1$ and close enough to 1, Table
\@ref(tab:theory) shows five theoretical predictions, along with our
corresponding empirical results for the correlation between polygenic scores and
RLRS.[^one-period-results] The key insight of the model is that for middling
levels of $\sigma$, the substitution effect dominates at low income levels, but
as income increases, the income and substitution effect balance out.

@@

Thus, a simple economic model can explain many of our results. Other empirical
work in economics also supports the link from human capital to fertility.
@caucutt2002women and @monstad2008education show that education and skills
affect age at first birth and fertility. Income decreases fertility at low
income levels, but increases it at higher income levels [@cohen2013financial].
US fertility decreases faster with education among single mothers than married
mothers [@baudin2015fertility], in line with our prediction 3 and as predicted
by @becker1981treatise. A related literature shows negative correlations between 
IQ and fertility [e.g. @lynn2004new; @reeve2018systematic].

@@

We test the economic theory in two ways. First, it predicts that
genetic variants will be selected for (or against) in proportion to their
correlation with human capital. Figure \@ref(fig:plot-cors-earnings-educ) plots
selection effects on each polygenic score against that score's correlation with
two measures of human capital: earnings in a respondent's first job, and
educational attainment. The relationships are strongly negative. Thus, human
capital appears to be relevant to natural selection. The negative relationship
suggests that substitution effects dominate income effects, which fits the known
negative association between income and fertility
[@becker1960economic;@jones2006economic]. The correlations reverse when we
control for age at first live birth, suggesting that within AFLB categories, the
income effect dominates.

@@

Second, we run a mediation analysis to directly test whether the correlation
between each polygenic score and fertility is mediated by educational attainment
(Appendix Table \@ref(tab:tbl-res-mediation)). We use the 
`r nrow(res_mediation)` scores where the selection effect is significant at $p$ <
0.05/33. Figure \@ref(fig:plot-mediation) shows estimated
proportions explained by educational attainment, along with bootstrap 95%
confidence intervals (uncorrected; 100 bootstraps). For `r n_pos_sig` scores,
the indirect effect of the score on fertility *via* educational attainment takes
the same sign as the overall effect, and is significantly different from zero
($p$ < 0.05/`r nrow(res_mediation)`). Among these scores, the median proportion of
the total effect explained by the indirect effect is 
`r scales::percent(median_prop)`. The educational attainment variable is a
relatively crude measure of human capital: more accurate measures would likely
explain more of the total effect.

```{r plot-mediation, fig.align = "center", fig.cap = "Proportion of selection effect mediated by educational attainment, among polygenic scores with significant selection effects"}

res_mediation %>% 
      mutate(
        term = pretty_names(term),
        term = fct_reorder(term, prop_ind, order_abs(1)),
        `Indirect effect p < 0.05/23` = sig 
      ) %>% 
      ggplot(aes(y = term)) + 
        geom_col(aes(x = prop_ind, fill = `Indirect effect p < 0.05/23`), 
                   width = .5) + 
        geom_errorbar(aes(
                          xmin = prop_ind_conf_low, 
                          xmax = prop_ind_conf_high,
                          linetype = "Bootstrap 95% c.i.",
                        ),
                        width = 0.3,
                        color = "grey50"
                      ) +
        scale_fill_manual(values = pval_palette) +
        my_vline +
        scale_x_continuous(labels = scales::percent, n.breaks = 7) +
        coord_cartesian(xlim = c(-0.5, 1)) +
        labs(x = "Proportion mediated", y = "") +
        scale_linetype_manual(name = "", values = c("Bootstrap 95% c.i." = 1))

@@

We consider three alternative theories that might explain our results. First,
welfare benefits which incentivize child-bearing might be taken up more among
low-income people. However, the majority of effect sizes appear unchanged
over a large span of twentieth-century history (Appendix Table
\@ref(tab:tbl-change-by-period)), during
which government spending on child-related benefits varied considerably
[@hoc1999socsec]. In general, there is only weak evidence that welfare benefits
affect fertility [@gauthier2007impact; see also @bergsvik2021can]. Future
work could test this theory more explicitly. A second alternative theory is that
polygenic scores correlate with the motivation to have children, i.e. parameter
$a$ in the model [cf. @jones2008fertility]. This theory would not explain why
selection effects are smaller at higher incomes and education levels. In fact,
in the model, $a$'s effect on fertility gets stronger at higher levels
of human capital. A third alternative is that traits under selection are linked
to externalizing behaviour and risk-seeking. This might be partially captured by
our parameter $\sigma$, which can be interpreted as a measure of risk aversion
over income; a more direct channel is risky sexual behaviour
[@mills2021identification]. The data here provide some support for this story: scores
which might plausibly be linked to externalizing behaviour, like ADHD and
younger age at smoking initiation, are selected for. However, risk-seeking 
seems unlikely to explain variation in fertility across the full range of scores
under selection, including physical measures like waist-hip ratio and BMI. We
test this theory directly by re-estimating equation \@ref(eq:regression)
controlling for a measure of risk attitude (UK Biobank field 2040). The median
ratio of effect sizes between regressions with and without controls is 
`r median(ratio)`; all scores which are significant at $p < 0.05/33$ in
uncontrolled regressions remain so when controlling for risk attitude. This
non-result could simply reflect the imprecision of the risk attitude measure,
which is a single yes/no question. But this measure *does* predict the overall
number of children, highly significantly ($p < 2 \times 10^{-16}$ in 
`r n_risk_sig` out of 33 regressions). Given that, and the statistical power we
get from our sample size, we believe that the non-result is real: while risk
attitude does predict fertility in the sample, it is not an important channel
for natural selection.

@@

To learn more about this, we compare effect sizes excluding and including
childless people in the *current* generation. The correlation between the two
sets of effect sizes is `r cor_inc_vs_excl_childless`. So, patterns across
different scores are broadly similar whether the childless are counted or not.
However, absolute effect sizes are smaller when the childless are excluded, for
`r n_smaller` out of 33 scores; the median percentage change is 
`r scales::percent(median_pct_diff)`.
